---
title: "Data Transformation"
author: "Oluwatobi Akinyemi - 3/16/2019"
date: "April 8, 2019"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Data Transformation

This file covers important code syntax to perform some of the most common functions for transforming data in R. It will also serve as a readily accessible reference for some hard to be remember functions which have great utility for the same.

### Filtering with filter()

Use filter to subset a data set by only returning rows that meet specific conditions.

```{r filter, echo=FALSE, warning=FALSE}
filter(dataframe, condition(s)) # Each condition can be separated by a comma sign
```

Creating a new dataframe using filter:

```{r filter, echo=FALSE, warning=FALSE}
newdf <- filter(df, condition(s))
```

Filter works with the usual comparison operators

You can also set multiple conditions to filter in the same column using boolean operators:

```{r filter, echo=FALSE, warning=FALSE}
newdf <- filter(df, col1 == n | col1 == m) # m & n are specific values
```

A great way to filter for matches of multiple values use `r %in%` as follows:
```{r filter, echo=FALSE, warning=FALSE}
filter(df, col1 %in% c(m, n, o, p))
```

An easy way to use filter to include or exclude many values that meet a broad category at once is to use the `r !` operator:
```{r filter, echo=FALSE, warning=FALSE}
filter(df, !(col1 > m | col2 > m))
```

Use `r between()` to filter values in a numeric vector that fall in a specific range:

```{r filter, echo=FALSE, warning=FALSE}
filter(df, between(col1, value1, value2)) # Where value1 and value2 are the low and high edges of the range.
```

To filter for missing values using filter, use `r is.na()` as follows:

```{r filter, echo=FALSE, warning=FALSE}
filter(df, is.na(col1))
```

### Arrange rows with arrange()

To arrange a data frame by descending values in a column:
```{r arrange, echo=FALSE, warning=FALSE}
arrange(df, desc(col1))
```

When sorting, missing values alway appear at the end.

### Select columns with select()

Select determines the order the columns should be returned

#### Special wildcards to use with Select()

The following woldcards are especially useful in selecting columns of interest when it is too cumbersome to list all desired columns out in one select statement

`r start_with()` selects all column names that start with the provided string:
```{r}
select(df, starts_with("xyz"))
```

`r ends_with()` selects all column names that end with the provided string:
```{r}
select(df, ends_with("xyz"))
```

`r contains()` selects all column names that contain the provided string:
```{r}
select(df, contains("xyz"))
```

`r matches()` selects all column names that matches a regular expression:
```{r}
select(df, matches("(.)\\1"))
```

`r everything()`: When your interest in the order that variables are listed in the result are limited to only a handful, you can use `r everything()` to include all remaining varaibles after specifying the order of the few you had in mind:
```{r}
select(df, col3, col4, col1, everything()) 
# The resulting data frame will have column 3 starting, followed by column 4, and column 1, then the remaining variables
```

### Mutate

Mutate adds new columns by applying functions to existing columns. Create new variables using existing variables:
```{r}
mutate(df,
  newvar1 = col1 - col3,
  newvar2 = col4 / col5 * 60
)
```

### Transmute

Use `r Transmute()` to drop old variables and only keep the newly created variables

%/% Integer division
%% Remainder

To break time into hours and minutes using modular arithmetic and transmute:

```{r}
transmute(flights,
  dep_time,
  hour = dep_time %/% 100,
  minute = dep_time %% 100
) 
```

### Summarize

Always precede `r summarise` with `r group_by`

Counting number of rows with summarise:
```{r}
by_dest <- group_by(flights, dest)
delay <- summarise(by_dest,
  count = n(),
  dist = mean(distance, na.rm = TRUE),
  delay = mean(arr_delay, na.rm = TRUE)
)
```

Counting using count():
```{r}
df %>% 
  count(col1)
```

