---
title: "Data Importing"
author: "Oluwatobi Akinyemi - 3/16/2019"
date: "April 5, 2019"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
```

## Data Import

This file covers important code syntax to perform some of the most common functions for importing data into R. It will also serve as a readily accessible reference for some hard to be remember functions which have great utility for the same.

### Saving and loading data

`r save()` is a great way to store multple data objects into one `r .RData` file as follows:

Given 3 data frames a, b and c

```{r}
save(a, b, c, file = "myobjects.RData")
```

To load the above saved file:
```{r}
load(file = "data/myobjects.RData")
```

### Clear workspace

To clear the global environment of all objects use:
```{r}
rm(list = ls())
```

### Reading in data

Reading in a .txt file uses the read.table function as follows:
```{r}
# Specify the location of the file using the file argument or if you set the working directory to the folder containing the file, you can simply use the file name
# \t tells r that the columns are tab-delimited for accurate reading
# Setting header = TRUE indicates that the first row of the data should be used as header
read.table(file = 'data/sample1.txt', 
           sep = '\t', 
           header = TRUE,
           stringsAsFactors = FALSE)    # do NOT convert strings to factors!!
```

We can read in data directly from a URL similarly by simply setting the `r file` argument of `r read.table` to the URL link to the desired file. Other arguments can be used as usual.

#### `r read_csv` vs `r read.csv`

When using `r read.csv`, some character columns are read in as factors, on the other hand `r read_csv` tends to keep such columns as character types. Also, `r read_csv`, which uses the `r readr` package tends to be a bit more efficient than `r read.csv`.

To read in a csv file use:
```{r}
# To read in sample1.csv
# Provided sample1 is in the working directory
sample1 <- read_csv("sample1.csv")
```

To read in semi-colon delimited files, use `r read_csv2()`
For tab-delimited files use `r read_tsv()`
For fixed-width files use `r read_fwf()`
For fixed-width files separated by white space use `r read_table()`
For files with any delimiter use `r read_delim`

To read a file sample2 whose columns are separated by "|" use:
```{r}
sample2 <- read_delim(sample2.txt, delim = "|")
```

To read excel files, use `r readxl`

### Exporting data

To export a data frame called sample1 to a .txt file use:
```{r}
# We can specify the desired file name using the file argument below
# \t makes the columns tab-delimited
write.table(x = sample1,
            file = "sample1.txt", sep = "\t")            
```

Writing to a csv:
```{r}
write_csv(dataframe, "filename.csv") # Where dataframe is data to be written and filename is file to be created
```

To retain column specifications use:
```{r}
write_rds(dataframe, "filename.rds") # Where dataframe is data to be written and filename is file to be created
```

### Parsing

```{r}
parse_logical() # parses logicals
parse_integer() # parses integers
parse_double() # parses double; numbers with decimal points
parse_number() # parses different kinds of numbers
parse_character() # parses characters
parse_factor() # parses factors
parse_date() # parses date
parse_datetime() # parses dates with a timestamp
parse_time() # parses time
```

To remove non numeric characters before or after a number use `r parse_number()` as follows:
```{r}
parse_number("$5000") # This will drop the '$' and return 5000
parse_number("It cost $123.45") # This will drop all non-numerics, the '$' and everything after the decimal point and return 123
```

Parsing numbers using the locale separator:
```{r}
# The local argument specifies the grouping mark used
parse_number("123.456.789", locale = locale(grouping_mark = "."))
```

To parse a string, you have to know the correct encoding the string is in.
To guess the encoding of a string use:
```{r}
guess_encoding(charToRaw())
```

You can use `r parse_factor()` to remove wrong values from a factor column with known levels as follows:
```{r}
# Specify levels using the levels argument of parse_factor
factor.levels <- c("level1", "level2")
parse_factor(c("column/list/strings"), levels = factor.levels)

fruit <- c("apple", "banana")
parse_factor(c("apple", "banana", "bananana"), levels = fruit)
```

Parse a date as follows:
```{r}
parse_datetime("20101010") # This will yield "2010-10-10 UTC"
```

parse_date wildcards:

Year
%Y (4 digits).
%y (2 digits);
Month
%m (2 digits).
%b (abbreviated, like "Jan").
%B (full, "January").
Day
%d (2 digits).
%e (optional leading space).
Time
%H 0-23 hour.
%I 0-12, must be used with %p.
%p AM/PM indicator.
%M minutes.
%S integer seconds.
%OS real seconds.
%Z Time zone (as name, e.g. America/Chicago). 
%z (as offset from UTC, e.g. +0800).
Non-digits
%. skips one non-digit character.
%* skips any number of non-digits.
