---
title: "Data Visualization"
author: "Oluwatobi Akinyemi - 3/16/2019"
date: "April 8, 2019"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Data Visualisation

### Creating a ggplot chart
data and mapping can be both placed in the ggplot statement as long as the mappings are the same for all geoms. Typically, they are separated as follows for flexibility:

```{r}
ggplot(data = mpg) + 
  geom_point(mapping = aes(x = displ, y = hwy))
```

The generic ggplot template:
```{r}
ggplot(data = <DATA>) + 
  <GEOM_FUNCTION>(mapping = aes(<MAPPINGS>))
```

### Editing the ggplot

You use aesthetic mappings to change the plot by mapping different variables to the geom function aesthetics. For example:
```{r}
ggplot(data = mpg) + 
  geom_point(mapping = aes(x = displ, y = hwy, color = class))
```

`r displ`, `r hwy` and `r color` are all variables in the `r mpg` dataframe. You can map different variables to different aesthetics such as shape, size, group to get the desired result.

The `r alpha` aesthetic is very useful in creating distinctions in the chart using various levels of transparency:
```{r}
ggplot(data = mpg) + 
  geom_point(mapping = aes(x = displ, y = hwy, alpha = class))
```

You can manually assign the value of a specific aesthetic, simply assign the desired value to the aesthetic otuside the `r aes()` as follows:
```{r}
ggplot(data = mpg) + 
  geom_point(mapping = aes(x = displ, y = hwy), color = "blue")
```

### Faceting

Use facets to subset the plots by a variable in the dataset. This works best with categorical or factor variables. Faceting with a continuous variable with yeild an output of unusable graphics. 
Use `r facet_wrap` to facet with a variable as follows:
```{r}
ggplot(data = mpg) + 
  geom_point(mapping = aes(x = displ, y = hwy)) + 
  facet_wrap(~ class, nrow = 2) # nrow specifies the number of rows in the facet grid
```

Faceting using multiple variables

You can facet with multiple variables:
```{r}
ggplot(data = mpg) + 
  geom_point(mapping = aes(x = displ, y = hwy)) + 
  facet_grid(drv ~ cyl) # Use ~ to combine both variables in the facet
```

### Geoms

Each geom has its own aesthetics that can be mapped to variables or assigned a value.

```{r}
# show.legend = FALSE turns off the legend
ggplot(data = mpg) +
  geom_smooth(
    mapping = aes(x = displ, y = hwy, color = drv),
    show.legend = FALSE
  )
```

Plot multiple geoms on the same plot:
```{r}
ggplot(data = mpg, mapping = aes(x = displ, y = hwy)) + 
  geom_point() + 
  geom_smooth()
```
By placing the axis mappings in the ggplot object, they will be carried forward in each geom function

### Statistical transformations

Many geom functions have a default stat associated with it.
See geom and stat pairs here: https://ggplot2.tidyverse.org/reference/
See ggplot2 cheat sheet here: https://www.rstudio.com/wp-content/uploads/2015/03/ggplot2-cheatsheet.pdf

Using stat_summary to plot a summary of the data:
```{r}
ggplot(data = diamonds) + 
  stat_summary(
    mapping = aes(x = cut, y = depth),
    fun.ymin = min,
    fun.ymax = max,
    fun.y = median
  )
```

### Color vs fill & Positions

Using the `r color` aesthetic controls the color of th eoutline of the object e.g. barchart. On the other hand, the `r fill` aesthetic controls the color of the bars themselves.

`r postion = "identity"` places each object in true location
`r postion = "fill"` stacks the groups in each bar, but keeps all bars at the same height
`r postion = "dodge"` places overlapping objects next to each other.

Use `r postion = "jitter"` or `r alpha` to address overplotting issues.

### Coordinate systems

To improve the visibility of axis labels use `r coord_flip()`
```{r}
ggplot(data = mpg, mapping = aes(x = class, y = hwy)) + 
  geom_boxplot() +
  coord_flip()
```
